---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mitochondriR

<!-- badges: start -->

[![R-CMD-check](https://github.com/selkamand/mitochondriR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/selkamand/mitochondriR/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/selkamand/mitochondriR/graph/badge.svg)](https://app.codecov.io/gh/selkamand/mitochondriR)
![GitHub Issues or Pull Requests](https://img.shields.io/github/issues-closed/selkamand/mitochondriR)
[![code size](https://img.shields.io/github/languages/code-size/selkamand/mitochondriR.svg)](https://github.com/selkamand/mitochondriR)
![GitHub last commit](https://img.shields.io/github/last-commit/selkamand/mitochondriR)
<!-- badges: end -->

> [!WARNING]  
> This package is in early development and not yet ready for use

mitochondriR estimates mitochondrial genomes per autosome ($\text{mtDNA}_\text{CN}$) from bulk whole-genome sequencing data using basic coverage information (mosdepth or idxstats file input) with optional correction for tumour aneuploidy using purity + ploidy.

## Calculation

The basic estimate is:

$$
\text{mtDNA}_\text{CN} = N_A \times \frac{C_M}{C_A}
$$

where:

$$
\begin{aligned}
C_A &= \text{median autosomal depth of coverage}, \\
C_M &= \text{median mitochondrial chromosome depth of coverage}, \\
N_A &= \text{average autosomal copy number per cell}.
\end{aligned}
$$
## Calculating mt genomes per cell 

Importantly, $\text{mtDNA}_\text{CN}$` does not necessarily reflect the number of
mitochondria per cell as a single mitochondrion can contain zero, one, or multiple mitochondrial genomes. 

However we can convert our mt genomes per autosome estimate into mt genomes per cell by simply multiplying $\text{mtDNA}_\text{CN}$` by number of autosomes per cell ($N_A$) once more: 

$$
\text{Mitochondrial Genomes Per Cell} = {N_A}^2 \times \frac{C_M}{C_A}
$$
This per-cell estimate is what the functions in this package output by default. Note that when you expect a large amount of error in your estimate of $N_A$, you might consider setting (`per_cell=FALSE`) to switch back to Mitochondral genomes per autosome as the squared ${N_A}$ coefficient of the per-cell metric can amplify these errors.

## Tumour purity and ploidy correction

In bulk tumour samples, the autosomal copy number may differ from the diploid ($N_A = 2$) profile we expect in healthy cells due to the mixture of tumour and normal cells and tendency of tumour cells to be aneuploid. If the tumour purity is
\(\rho\), and the tumour ploidy is \(P\), then the typical autosomal copy number
in the bulk sample is estimated as:

$$
N_A = 2(1 - \rho) + \rho P
$$

where:

$$
\begin{aligned}
\rho &= \text{tumour purity}, \\
P &= \text{tumour ploidy}.
\end{aligned}
$$

For example, in a sample with 80% tumour purity and tumour ploidy 3:

$$
\begin{aligned}
N_A &= 2(1 - 0.8) + 0.8 \times 3 \\
    &= 0.4 + 2.4 \\
    &= 2.8.
\end{aligned}
$$

The purity- and ploidy-adjusted estimate is therefore:

$$
\text{mtDNA}_\text{CN} = \left[2(1 - \rho) + \rho P\right] \times \frac{C_M}{C_A}
$$

where:

$$
\begin{aligned}
\rho &= \text{tumour purity}, \\
P &= \text{tumour ploidy}. \\
C_A &= \text{median autosomal depth of coverage}, \\
C_M &= \text{median mitochondrial chromosome depth coverage}, \\
N_A &= \text{average autosomal copy number per cell}.
\end{aligned}
$$

## Other notes 

The quality of our estimate of 'median coverage' for autosomes and mitochondrial genomes will depend on what input file you start from. 
Samtools idxstats files are fast and convenient but only allow for average depth of coverage to be computed for any one chromosome. 
For autosomes we can take the median number of mapped reads per KB (to mitigate impact of aneuploid autosomes) but this metric will still be inflated by high depth alignments to short repetitive regions. 
Similarly, mitochondrial genome coverage can only be calculated as an average. 

This is why we highly recommend mosdepth per autosomal base (or region) based median depth of coverage estimates. Idxstats support is only really provided for convenience.

## Installation

You can install the development version of mitochondriR like so:

``` r
if (!require("remotes"))
    install.packages("remotes")

remotes::install_github("selkamand/mitochondriR")
```

## Quick Start

### From idxstats file (not recommended) 

See 'Other notes' section above for why we recommend mosdepth-based estimation.

Ensure you've run `samtools idxstats` on your alignment to a reference genome.

```{r idxstats}
library(mitochondriR)

idxstats <- system.file("idxstats", package = "mitochondriR")

# Compute rouch mtDNA copynumber from idxstats file, assuming 150bp read length
mtdna_from_idxstats(
  idxstats,
  background_ploidy = 2,
  tumour_ploidy = 2, 
  tumour_purity = 1
)

```

### Calculation Only

If you estimate each of the required parameters using another method, you can perform the mtDNAcn calculation below:

```{r manual_calculation}
library(mitochondriR)

calculate_mtdna_tumour(
  mitochondrial_depth = 100,
  autosome_depth = 10,
  background_ploidy = 2,
  tumour_ploidy = 2, 
  tumour_purity = 1
)
```

Or if your not sequencing a tumour sample and don't have to worry about mixed ploidy, you can use the simpler interface:

```{r}
calculate_mtdna(
  mitochondrial_depth = 100,
  autosome_depth = 10,
  ploidy = 2
)
```

